Metric Spaces with Expensive Distances

### 2019-03-30 paper/draft_esa.tex
Interim for ESA

\documentclass[a4paper,USenglish]{socg-lipics-v2018}
%\documentclass[a4paper,USenglish]{article}

%\usepackage{arxiv}

@@

\title{Metric Spaces with Expensive Distances}
\author{Michael Kerber}{TU Graz}{kerber@tugraz.at}{https://orcid.org/0000-0002-8030-9299}{Supported by Austrian Science Fund (FWF) grant number P 29984-N35}
\author{Arnur Nigmetov}{TU Graz}{nigmetov@tugraz.at}{}{}
\authorrunning{M. Kerber and A. Nigmetov}

\keywords{metric spaces, doubling dimension, spanners, approximate nearest neighbor}
%\subjclass{\ccsdesc[500]{Theory of computation~Nearest neighbor algorithms}}
\subjclass{\ccsdesc[500]{Theory of computation~Random search heuristics}}
% \category{dummy}

\Copyright{Michael Kerber and Arnur Nigmetov}

@@

@@

@@

@@

\definecolor{uququq}{rgb}{0.25,0.25,0.25}
\definecolor{xdxdff}{rgb}{0.49,0.49,1}
\definecolor{qqqqff}{rgb}{0,0,1}
\begin{figure}[ht!]
    \centering
\begin{tikzpicture}[line cap=round,line join=round,>=triangle 45,x=0.5cm,y=0.5cm]
%\draw[->,color=black] (-1.5,0) -- (17,0);
%\foreach \x in {,2,4,6,8,10,12,14,16}
%\draw[shift={(\x,0)},color=black] (0pt,2pt) -- (0pt,-2pt) node[below] {\footnotesize $\x$};
%\draw[->,color=black] (0,-7.5) -- (0,9.2);
%\foreach \y in {-6,-4,-2,2,4,6,8}
%\draw[shift={(0,\y)},color=black] (2pt,0pt) -- (-2pt,0pt) node[left] {\footnotesize $\y$};
%\draw[color=black] (0pt,-10pt) node[right] {\footnotesize $0$};
\clip(-1.5,-7.5) rectangle (17,9.2);
\draw [fill=black,fill opacity=0.5] (8.52,0.98) circle (1.02cm);
\draw [fill=black,fill opacity=0.15] (8.52,0.98) circle (4cm);
\draw [fill=black,fill opacity=0.95] (6.12,3.39) circle (0.42cm);
\draw [fill=black,fill opacity=0.5] (7.65,-3.14) circle (1.32cm);
\draw [line width=2pt] (3.51,0.63) circle (1.9cm);
\draw [dash pattern=on 2pt off 2pt] (3.51,0.63) circle (1.49cm);
\draw (3.51,0.63)-- (0.62,1.33);
\draw (3.51,0.63)-- (6.12,3.39);
\draw (8.52,0.98)-- (3.51,0.63);
\draw (8.52,0.98)-- (14.89,-3.85);
\draw (11.88,-1) node[anchor=north west] {$r_2 +r_1 / ( 1 + \eps$};
\begin{scriptsize}
\fill [color=qqqqff] (3.51,0.63) circle (1.5pt);
\draw[color=qqqqff] (3.71,0.94) node {$q$};
\fill [color=xdxdff] (0.62,1.33) circle (1.5pt);
\draw[color=black] (3.24,1.41) node {$r_1/(1+\eps)$};
\fill [color=qqqqff] (8.52,0.98) circle (1.5pt);
\draw[color=qqqqff] (8.9,1.29) node {$p_2$};
\fill [color=uququq] (6.48,0.83) circle (1.5pt);
\fill [color=qqqqff] (7.83,2.21) circle (1.5pt);
\draw[color=qqqqff] (8.21,2.52) node {$p_3$};
\fill [color=qqqqff] (7.65,-3.14) circle (1.5pt);
\draw[color=qqqqff] (8.01,-2.82) node {$p_4$};
\fill [color=qqqqff] (6.62,0.6) circle (1.5pt);
\draw[color=qqqqff] (6.98,0.92) node {$p_5$};
\fill [color=qqqqff] (1.49,-2.39) circle (1.5pt);
\draw[color=qqqqff] (1.87,-2.08) node {$p_6$};
\fill [color=qqqqff] (1.95,-2.65) circle (1.5pt);
\draw[color=qqqqff] (2.31,-2.33) node {$p_7$};
\fill [color=qqqqff] (2.3,-2.75) circle (1.5pt);
\draw[color=qqqqff] (2.68,-2.43) node {$p_8$};
\fill [color=qqqqff] (2.79,-2.9) circle (1.5pt);
\draw[color=qqqqff] (3.17,-2.57) node {$p_9$};
\fill [color=xdxdff] (14.89,-3.85) circle (1.5pt);
\fill [color=uququq] (5.84,3.09) circle (1.5pt);
\fill [color=xdxdff] (4.52,4.29) circle (1.5pt);
\draw[color=xdxdff] (4.89,4.61) node {$p_1$};
\end{scriptsize}
\end{tikzpicture}
\caption{First two steps of the ANN algorithm. First $p_1$ is 
chosen as the current candidate, and we must compute $\dist(p_2, q)$. After that the algorithm will not compute
distance to any of the points inside the heavily shaded ball or outside the lightly shaded ball that are centered at $p_2$, because their lower bounds allow us to discard them. Note that the point $p_5$, which
is closer to $q$ than $p_1$, also will not be a candidate, and at least one of the points $p_6,p_7,p_8,p_9$ in the annulus between the dashed and solid circle, which are further from $q$ than $p_5$, will be chosen as $c$. This shows that in our algorithm the distance from the candidate to $q$ can drop \textit{slower} than in the bruteforce algorithm, thus Theorem \ref{thm:ann_bound} does not immediately follow from standard backwards analysis. The small black ball between the dashed circle and the solid circle has radius $v_1 \eps / (1 + \eps)$; it is the ball that we use in the packing argument, because it is smaller than any of the lightly shaded balls that correspond to points like $p_2$ and $p_4$, that is, the points that do not improve $v$.}
\label{fig:ann_illustration}
\end{figure}

@@

\bibliography{bib}
\bibliographystyle{plain}

>>>>

\documentclass[a4paper,UKenglish,cleveref, autoref]{lipics-v2019}
%\documentclass[a4paper,USenglish]{article}

\bibliographystyle{plainurl}% the mandatory bibstyle

@@

\title{Metric Spaces with Expensive Distances}
\titlerunning{Metric Spaces with Expensive Distances}
% uncomment for LIPICS 
\author{Michael Kerber}{TU Graz}{kerber@tugraz.at}{https://orcid.org/0000-0002-8030-9299}{Supported by Austrian Science Fund (FWF) grant number P 29984-N35}
\author{Arnur Nigmetov}{TU Graz}{nigmetov@tugraz.at}{}{}
\authorrunning{M. Kerber and A. Nigmetov}

\keywords{metric spaces, doubling dimension, spanners, approximate nearest neighbor}
\subjclass{\ccsdesc[500]{Theory of computation~Random search heuristics}}
% \category{dummy}

\Copyright{Michael Kerber and Arnur Nigmetov}
% END uncomment for LIPICS

% uncomment for Arxiv
%\author{Michael Kerber\\
%Graz University of Technology\\
%Institut fuer Geometrie\\
%Kopernikusgasse 24\\
%8010 Graz\\
%Austria\\
%kerber@tugraz.at\\
%Supported by Austrian Science Fund (FWF) grant number P 29984-N35
%\And

%Arnur Nigmetov
%Graz University of Technology\\
%Institut fuer Geometrie\\
%Kopernikusgasse 24\\
%8010 Graz\\
%Austria\\
%nigmetov@tugraz.at}

%\keywords{metric spaces, doubling dimension, spanners, approximate nearest neighbor}

% END uncomment for arxiv

@@

% uncomment for Arxiv
%\keywords{metric spaces, doubling dimension, spanners, approximate nearest neighbor}
% END uncomment for Arxiv

@@

\begin{figure}[ht]
    \begin{centering}
\begin{tikzpicture}
    \begin{axis}[xlabel=\# Diagrams,ylabel=\# Edges / \# Diagrams,
        legend pos=outer north east]

    %\addplot+ table [x=n_points, y=ratio_greedy_dim_0_eps_01_q_1, col sep=comma] {mcgill_perturbed_blind_greedy.csv};
    %\addlegendentry{Greedy, $\eps = 0.1$}

    %\addplot+ table [x=n_points, y=ratio_blind_greedy_dim_0_eps_01_q_1, col sep=comma] {mcgill_perturbed_blind_greedy.csv};
    %\addlegendentry{Blind greedy, $\eps = 0.1$}

    \addplot+ table [x=n_points, y=ratio_greedy_dim_0_eps_02_q_1, col sep=comma] {mcgill_perturbed_blind_greedy.csv};
    \addlegendentry{Greedy, $\eps = 0.2$}

    \addplot+ table [x=n_points, y=ratio_blind_greedy_dim_0_eps_02_q_1, col sep=comma] {mcgill_perturbed_blind_greedy.csv};
    \addlegendentry{Blind greedy, $\eps = 0.2$}

    \addplot+ table [x=n_points, y=ratio_greedy_dim_0_eps_05_q_1, col sep=comma] {mcgill_perturbed_blind_greedy.csv};
    \addlegendentry{Greedy, $\eps = 0.5$}

    \addplot+ table [x=n_points, y=ratio_blind_greedy_dim_0_eps_05_q_1, col sep=comma] {mcgill_perturbed_blind_greedy.csv};
    \addlegendentry{Blind greedy, $\eps = 0.5$}

\end{axis}
\end{tikzpicture}
\end{centering}
    %\includegraphics[width=0.7\textwidth]{pics/edges_to_points_ratio_dim_2_normal_points.png}
    \caption{Ratio \# edges / \# diagrams for blind greedy spanner 
    algorithm. The plot is for 0-dimensional diagrams of perturbed McGill data
    $\eps = 0.1, 0.2, 0.5$ with $W_1$ distance.}
    \label{fig:spanner_ratio_mcgill_perturbed}
\end{figure}

\begin{figure}[ht]
    \begin{centering}
\begin{tikzpicture}
    \begin{axis}[xlabel=\# Diagrams,ylabel=\# Edges / \# All distances,
        legend pos=outer north east]

    \addplot+ table [x=n_points, y=sparseness_greedy_dim_0_eps_01_q_1, col sep=comma] {mcgill_perturbed_blind_greedy.csv};
    \addlegendentry{Greedy, $\eps = 0.1$}

    \addplot+ table [x=n_points, y=sparseness_blind_greedy_dim_0_eps_01_q_1, col sep=comma] {mcgill_perturbed_blind_greedy.csv};
    \addlegendentry{Blind greedy, $\eps = 0.1$}

    \addplot+ table [x=n_points, y=sparseness_greedy_dim_0_eps_02_q_1, col sep=comma] {mcgill_perturbed_blind_greedy.csv};
    \addlegendentry{Greedy, $\eps = 0.2$}

    \addplot+ table [x=n_points, y=sparseness_blind_greedy_dim_0_eps_02_q_1, col sep=comma] {mcgill_perturbed_blind_greedy.csv};
    \addlegendentry{Blind greedy, $\eps = 0.2$}

    \addplot+ table [x=n_points, y=sparseness_greedy_dim_0_eps_05_q_1, col sep=comma] {mcgill_perturbed_blind_greedy.csv};
    \addlegendentry{Greedy, $\eps = 0.5$}

    \addplot+ table [x=n_points, y=sparseness_blind_greedy_dim_0_eps_05_q_1, col sep=comma] {mcgill_perturbed_blind_greedy.csv};
    \addlegendentry{Blind greedy, $\eps = 0.5$}

\end{axis}
\end{tikzpicture}
\end{centering}
    %\includegraphics[width=0.7\textwidth]{pics/edges_to_points_ratio_dim_2_normal_points.png}
    \caption{Ratio \# edges / \# all distances for blind greedy spanner 
    algorithm. The plot is for 0-dimensional diagrams of perturbed McGill data
    $\eps = 0.1, 0.2, 0.5$ with $W_1$ distance.}
    \label{fig:spanner_ratio}
\end{figure}

@@

We also ran experiments on a more realistic data set.

We also generate real persistence diagrams of the shapes from  the McGill shape benchmark \cite{zhang2005mcgill}, 
to which we add some random noise.

@@

\definecolor{uququq}{rgb}{0.25,0.25,0.25}
\definecolor{xdxdff}{rgb}{0.49,0.49,1}
\definecolor{qqqqff}{rgb}{0,0,1}
\begin{figure}[ht!]
    \centering
\begin{tikzpicture}[line cap=round,line join=round,>=triangle 45,x=0.5cm,y=0.5cm]
%\draw[->,color=black] (-1.5,0) -- (17,0);
%\foreach \x in {,2,4,6,8,10,12,14,16}
%\draw[shift={(\x,0)},color=black] (0pt,2pt) -- (0pt,-2pt) node[below] {\footnotesize $\x$};
%\draw[->,color=black] (0,-7.5) -- (0,9.2);
%\foreach \y in {-6,-4,-2,2,4,6,8}
%\draw[shift={(0,\y)},color=black] (2pt,0pt) -- (-2pt,0pt) node[left] {\footnotesize $\y$};
%\draw[color=black] (0pt,-10pt) node[right] {\footnotesize $0$};
\clip(-1.5,-7.5) rectangle (17,9.2);
\draw [fill=black,fill opacity=0.5] (8.52,0.98) circle (1.02cm);
\draw [fill=black,fill opacity=0.15] (8.52,0.98) circle (4cm);
\draw [fill=black,fill opacity=0.95] (6.12,3.39) circle (0.42cm);
\draw [fill=black,fill opacity=0.5] (7.65,-3.14) circle (1.32cm);
\draw [line width=2pt] (3.51,0.63) circle (1.9cm);
\draw [dash pattern=on 2pt off 2pt] (3.51,0.63) circle (1.49cm);
\draw (3.51,0.63)-- (0.62,1.33);
\draw (3.51,0.63)-- (6.12,3.39);
\draw (8.52,0.98)-- (3.51,0.63);
\draw (8.52,0.98)-- (14.89,-3.85);
\draw (11.88,-1) node[anchor=north west] {$r_2 +r_1 / ( 1 + \eps)$};
\begin{scriptsize}
\fill [color=qqqqff] (3.51,0.63) circle (1.5pt);
\draw[color=qqqqff] (3.71,0.94) node {$q$};
\fill [color=xdxdff] (0.62,1.33) circle (1.5pt);
\draw[color=black] (3.24,1.41) node {$r_1/(1+\eps)$};
\fill [color=qqqqff] (8.52,0.98) circle (1.5pt);
\draw[color=qqqqff] (8.9,1.29) node {$p_2$};
\fill [color=uququq] (6.48,0.83) circle (1.5pt);
\fill [color=qqqqff] (7.83,2.21) circle (1.5pt);
\draw[color=qqqqff] (8.21,2.52) node {$p_3$};
\fill [color=qqqqff] (7.65,-3.14) circle (1.5pt);
\draw[color=qqqqff] (8.01,-2.82) node {$p_4$};
\fill [color=qqqqff] (6.62,0.6) circle (1.5pt);
\draw[color=qqqqff] (6.98,0.92) node {$p_5$};
\fill [color=qqqqff] (1.49,-2.39) circle (1.5pt);
\draw[color=qqqqff] (1.87,-2.08) node {$p_6$};
\fill [color=qqqqff] (1.95,-2.65) circle (1.5pt);
\draw[color=qqqqff] (2.31,-2.33) node {$p_7$};
\fill [color=qqqqff] (2.3,-2.75) circle (1.5pt);
\draw[color=qqqqff] (2.68,-2.43) node {$p_8$};
\fill [color=qqqqff] (2.79,-2.9) circle (1.5pt);
\draw[color=qqqqff] (3.17,-2.57) node {$p_9$};
\fill [color=xdxdff] (14.89,-3.85) circle (1.5pt);
\fill [color=uququq] (5.84,3.09) circle (1.5pt);
\fill [color=xdxdff] (4.52,4.29) circle (1.5pt);
\draw[color=xdxdff] (4.89,4.61) node {$p_1$};
\end{scriptsize}
\end{tikzpicture}
\caption{First two steps of the ANN algorithm. First $p_1$ is 
chosen as the current candidate, and we must compute $\dist(p_2, q)$. After that the algorithm will not compute
distance to any of the points inside the heavily shaded ball or outside the lightly shaded ball that are centered at $p_2$, because their lower bounds allow us to discard them. Note that the point $p_5$, which
is closer to $q$ than $p_1$, also will not be a candidate, and at least one of the points $p_6,p_7,p_8,p_9$ in the annulus between the dashed and solid circle, which are further from $q$ than $p_5$, will be chosen as $c$. This shows that in our algorithm the distance from the candidate to $q$ can drop \textit{slower} than in the bruteforce algorithm, thus Theorem \ref{thm:ann_bound} does not immediately follow from standard backwards analysis. The small black ball between the dashed circle and the solid circle has radius $v_1 \eps / (1 + \eps)$; it is the ball that we use in the packing argument, because it is smaller than any of the lightly shaded balls that correspond to points like $p_2$ and $p_4$, that is, the points that do not improve $v$.}
\label{fig:ann_illustration}
\end{figure}

@@

\bibliography{bib}
